---
title: "Homework 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,eval=T,message=F,warning=F,fig.align='center')
```

Each part of each question will be 2.5pts, there are 20 parts, so 50pts total.


## 1) Catamaran, revisited

Startup pet supply company Catamaran is trying to better understand the spending behavior of its customers.
In particular, the company wants to find simple ways to predict how much customers will spend on Catamaran products from their purchases of just one such product: cat litter.

A (sadly, fictional) data set is stored in the file `catamaran.csv`, available from [here](https://pages.stat.wisc.edu/~bwu62/catamaran.csv).
Download this file and save it in the same directory as your working directory (you can check this directory with `getwd()`).
The data frame encoded in this file stores two columns:

1. The column titled `litter` is the amount of cat litter, in pounds, purchased by a customer in the past year (you'll see in the data that Catamaran sells litter in three-pound increments; no, I don't think that's a realistic increment in which to sell cat littler. Fictional data is fun!).
2. The column titled `spending` is the amount of money, in dollars, that a customer has spent on Catamaran products (including cat litter) in the past year.

The following block of code loads the data in this file into a data frame `catamaran`.

```{r}
catamaran = read.csv('https://pages.stat.wisc.edu/~bwu62/catamaran.csv')
```

### Part a) inspecting the data

Create a scatterplot showing customer spending as a function of how much cat litter they bought.
Do you see a linear trend?
Based just on looking at the scatterplot, what do you estimate the slope to be (you will not be graded on the accuracy of this estimate-- just give a best guess for fun to see how close it is to the estimated model!).

```{r}
library(ggplot2)
ggplot(catamaran, aes(x=litter, y=spending)) + geom_point() + theme_minimal() + labs(title = "Scatterplot of Spending vs Cat Litter", x = "Cat Litter (lbs)", y = "Spending ($)")
```

***

Based on the scatterplot, there appears to be a linear trend. A rough estimate of the slope could be around 15.

***

### Part b) fitting a model

Fit a linear model to the Catamaran data, regressing spending against the amount of litter purchased (and an intercept term).

Store the estimated intercept in a variable called `cat_intercept_hat`, and store the estimated coefficient of `litter` in a variable called `cat_slope_hat`.
Don't forget to use the `unname()` function to strip the labels off of these, ensuring that these two variables just store numbers.

```{r}
cat_model = lm(spending ~ litter, data = catamaran)
cat_intercept_hat = unname(coef(cat_model)[1])
cat_slope_hat = unname(coef(cat_model)[2])
cat_intercept_hat
cat_slope_hat
```

### Part c) interpreting the model

Based on these estimates, the purchase of one additional pound of cat litter per year is associated with how many more dollars per year spent on Catamaran products?

***

One additional pound of cat litter per year is associated with about cat_slope_hat more dollars spent on Catamaran products per year.

***

As we mentioned above, Catamaran sells cat littler in three-pound units.
Thus, a more natural question is: the purchase of one additional three-pound unit (i.e., three additional pounds) of cat littler is associated with an increase of how many more dollars per year spent on Catamaran products?

***

One additional three-pound unit of cat litter is associated with about 3 * cat_slope_hat more dollars spent on Catamaran products per year.

***

Perhaps a more sane increment in which to sell cat litter would be twenty-pound bags.
Based on your estimated coefficients, an additional twenty pounds of cat litter purchased per year is associated with an increase of how many more dollars per year spent on Catamaran products?

***

An additional twenty pounds of cat litter purchased per year is associated with about 20 * cat_slope_hat more dollars spent on Catamaran products per year.

***

### Part d) generating a confidence interval

Of course, Catamaran's data is noisy, so there is uncertainty in our estimate of the coefficients in our model.

Create a Q-Q plot to verify that the residuals of our model are approximately normal.
Do you see anything unusual?
You probably won't-- the observation errors in this fake data really are normal.
Still, take a look just to be sure; it's a good habit to always at least briefly check the appropriateness of your model.

```{r}
library(car)
qqPlot(residuals(cat_model), main = "Q-Q Plot of Model Residuals")
```

Once you've verified that the residuals look reasonable, and hence our normality assumptions are defensible, construct a 95% confidence interval for the coefficient of `litter` in our model.


```{r}
conf_int = confint(cat_model, level = 0.95)
conf_int
```

Based on this confidence interval, should we accept or reject the null hypothesis that $\beta_1=0$ at level $\alpha=0.05$?

***

Based on the 95% confidence interval, we should accept the null hypothesis that 𝛽1=0 at level 𝛼=0.05.

***

Finally, verify your answer by looking at the `summary` output of your model and check that the coefficient is or is not statistically significantly different from zero.

```{r}
summary(cat_model)
```


<br/><br/>

## 2) Understanding the effect of noise

This problem, loosely based on Problem 13 in Chapter 3 of [ISLR](https://www.statlearning.com/), will help to give you an intuition to the role of sample size (i.e., number of observations $n$) and  noise level (as captured by the variance $\sigma^2$ of the noise terms $\epsilon_i$).

### Part a) generating linear data

Write a function `generate_linear_data` that takes two arguments: `n` and `sigma2`, in that order, and does the following:

1. Use the `rnorm()` function to create a vector `x`, containing `n` independent observations drawn from a normal distribution with mean $0$ and variance $1$. This will represent our vector of predictors.

2. Use the `rnorm()` function to create a vector, `eps`, containing `n` independent observations drawn from a normal distribution with mean $0$ and variance `sigma2`. These will correspond to the errors in our observed responses.

3. Using `x` and `eps`, construct a vector `y` according to the model
$$
Y = -1 + 0.5X + \epsilon,
$$

where $X$ corresponds to entries in our vector `x` and $\epsilon$ corresponds to entries in our vector `eps`.

4. Create a data frame with two columns, `predictors` and `responses` whose entries correspond to the vectors `x` and `y`, respectively. Return this data frame.

You do not need to perform any error checking in this function.
You may assume that `n` is a positive integer and `eps` is a positive numeric.

Before writing code, let's __check your understanding:__ What is the length of the vector `y`? What are the values of the intercept $\beta_0$ and slope $\beta_1$ in this linear model?

***
The length of the vector y is equal to n. The values of the intercept 𝛽0 and slope 𝛽1 in this linear model are -1 and 0.5, respectively.
***

```{r}
generate_linear_data = function(n, sigma2) {
  x = rnorm(n, mean = 0, sd = sqrt(1))
  eps = rnorm(n, mean = 0, sd = sqrt(sigma2))
  y = -1 + 0.5 * x + eps
  data.frame(predictors = x, responses = y)
}
```

### Part b) Plotting data

Use your function from Part (a) to generate 100 samples from the model
$$
Y = -1 + 0.5X + \epsilon,
$$

with `sigma2` set to $0.25$ and create a scatterplot of that data, showing the responses $Y$ as a function of $X$.
You may use either `ggplot2` or R's built-in plotting utilities.

Examine the point cloud and discuss:
Does the data look approximately linear?
Does the slope look about right?
What about the intercept?
__Note:__ You __do not__ need to fit a model, yet! Just inspect the data!

```{r}
data1 = generate_linear_data(100, 0.25)
ggplot(data1, aes(x=predictors, y=responses)) + geom_point() + theme_minimal() + labs(title = "Scatterplot of Responses vs Predictors", x = "Predictors", y = "Responses")
```

***

The data looks approximately linear, the slope seems close to 0.5, and the intercept seems close to -1.

***

### Part c) the effect of noise

Now, generate 100 data points again, as in part (b), but increase the noise level (i.e., the variance of the observation errors $\epsilon$) to $1$.
That is, set `sigma2` to `1`.
Plot the data again, and compare to the previous plot.
What do you observe?

```{r}
data2 = generate_linear_data(100, 1)
ggplot(data2, aes(x=predictors, y=responses)) + geom_point() + theme_minimal() + labs(title = "Scatterplot of Responses vs Predictors (sigma2 = 1)", x = "Predictors", y = "Responses")
```

***

With a higher noise level, the data looks less linear and it is more difficult to discern the slope and intercept.

***

Now, try decreasing the noise level (i.e., the variance of the $\epsilon$ terms), down to $\sigma^2 = 0.1$ and create one more plot, again with $n=100$ data points.
What do you observe?

```{r}
data3 = generate_linear_data(100, 0.1)
ggplot(data3, aes(x=predictors, y=responses)) + geom_point() + theme_minimal() + labs(title = "Scatterplot of Responses vs Predictors (sigma2 = 0.1)", x = "Predictors", y = "Responses")
```

***

With a lower noise level, the data looks more linear and it is easier to discern the slope and intercept.

***

### Part d) estimating from synthetic data

Now, let's investigate how the amount of noise (i.e., the error term variance $\sigma^2$) influences our estimation of the slope $\beta_1$.
Hopefully in your plots above you noticed that when the variance $\sigma^2$ is larger, the linear trend in the data is "harder to see".
Perhaps unsurprisingly, but still interestingly, this translates directly into difficulty in estimating the coefficients.
When there is more noise in our observations, our estimation of the coefficients suffers.

Let's investigate this with a simulation. This part of the problem will have you write code to run a single experiment wherein we generate data and try to estimate the slope $\beta_1$.
In Part (e) below, we'll use this single-trial code to run a Monte Carlo simulation that estimates the variance of our estimate $\hat{\beta}_1$.
We'll be able to see how the variance of our estimate (i.e., how close we are on average to the true $\beta_1$) changes as the noise $\sigma^2$ changes.

Write a function `generate_and_estimate` that takes two arguments: a sample size `n` and a variance term `sigma2`, and does the following:

1. Use `generate_linear_data` to generate a collection of `n` observations from a linear model
$$
Y = -1 + 0.5X + \epsilon,
$$
where the noise term $\epsilon$ is normal with variance `sigma2`.

2. Pass this data into `lm()` to fit a model predicting the column `responses` from the column `predictors` and an intercept term.

3. Extract the estimate of the slope from the resulting fitted model object (hint: look at the `coefficients` attribute of the model object or use the function `coef()`). Call this `beta1hat`. __Hint:__ don't forget to use `unname()` to remove the "names" of the coefficients extracted from the model object.

4. Return `beta1hat`.

```{r}
generate_and_estimate = function(n, sigma2) {
  data = generate_linear_data(n, sigma2)
  model = lm(responses ~ predictors, data = data)
  beta1hat = unname(coef(model)[2])
  beta1hat
}
```

### Part e) estimating variance of an estimator

Now, let's write code compute a Monte Carlo estimate of the variance of our estimator $\hat{\beta}_1$.
Note that this variance is a good way to measure the (average) squared error of our estimator. When this variance is large, it means that our estimate of $\beta_1$ is more uncertain, as we expect to be farther from the true value of $\beta_1$ more often, on average.

Write a function `estimate_beta1hat_variance` that takes three arguments: a number of observations `n`, a variance `sigma2` and a number of Monte Carlo replicates `M`, and does the following:

1. Use `generate_and_estimate` to generate a collection of `n` observations from a linear model
$$
Y = -1 + 0.5X + \epsilon,
$$
where the noise term $\epsilon$ is normal with variance `sigma2`, and estimate $\beta_1$. Call the resulting estimate `beta1hat`.

2. Perform step 1 a total of `M` times, recording the resulting `beta1hat` each time in a vector. That is, perform `M` Monte Carlo iterations of the experiment wherein we generate random data and estimate the slope $\beta_1 = 0.5$, keeping track of our estimate in each Monte Carlo replicate.

3. Compute and return the variance of our `M` random `beta1hat` replicates. This is a Monte Carlo estimate of the variance of our estimate $\hat{\beta}_1$.
You may use either the corrected or uncorrected sample variance in this calculation.

```{r}
estimate_beta1hat_variance = function(n, sigma2, M) {
  beta1hats = replicate(M, generate_and_estimate(n, sigma2))
  var(beta1hats)
}
```

### Part f) effect of noise on estimation accuracy

Use your function from Part (e) to create a plot of the variance (as estimated from 1000 Monte Carlo iterates) of the estimator $\hat{\beta}_1$, as a function of $\sigma^2$, when $n=100$.
Use values for $\sigma^2$ ranging from $0.25$ to $4$, inclusive, in increments of $0.25$.
You may use either `ggplot2` or the built-in R plotting functions.

__Note:__ this simulation make take a few minutes to run, since for each value of $\sigma^2$, we must perform $M=1000$ simulations, and each simulation requires fitting linear regression, which is not free!

```{r}
sigma2_values = seq(0.25, 4, 0.25)
var_estimates = sapply(sigma2_values, function(s) estimate_beta1hat_variance(100, s, 1000))
df_var_estimates = data.frame(sigma2 = sigma2_values, var_estimates = var_estimates)
ggplot(df_var_estimates, aes(x=sigma2, y=var_estimates)) + geom_line() + theme_minimal() + labs(title = "Variance of Beta1hat vs Sigma2", x = "Sigma2", y = "Variance of Beta1hat")
```

Based on your plot, how does it look like the variance of our estimator $\hat{\beta}_1$ behaves as a function of the observation error variance $\sigma^2$?

If you look up the variance of $\hat{\beta}_1$ in a mathematical statistics textbook, you will find that
$$
\operatorname{Var} \hat{\beta}_1
=
\frac{ \sigma^2 }{ \sum_{i=1}^n (x_i - \bar{x})^2 }.
$$

Does this agree with your plot above?

***

Yes I agree because based on the plot, the variance of our estimator 𝛽̂ 1 increases as the observation error variance 𝜎2 increases.

***


<br/><br/>

## 3) More regression with `mtcars`

In lecture, we worked briefly with the `mtcars` data set.
Let's get more regression practice by working with it some more.

### a) background

Run `?mtcars` in the console (please __do not__ add it to this `Rmd` file) and briefly read the help page.
Specifically, take note of the following:

1. What is the source of this data?
2. What is this data set measuring (i.e., what was the response variable in the original study, at least based on the brief description in the R documentation)?
3. What predictors are available and what do they mean?

***

1. The source of the mtcars dataset is the 1974 Motor Trend US magazine.
2. The dataset measures the fuel consumption (miles per gallon, or mpg) and 10 other aspects of automobile design and performance.
3. The available predictors are:
cyl: number of cylinders
disp: displacement (cu.in.)
hp: gross horsepower
drat: rear axle ratio
wt: weight (1000 lbs)
qsec: 1/4 mile time
vs: engine (0 = V-shaped, 1 = straight)
am: transmission (0 = automatic, 1 = manual)
gear: number of forward gears
carb: number of carburetors

***

You may want to also run `head(mtcars, 10)` or `View(mtcars)` to inspect the data frame briefly before moving on.

### b) Fitting a model

Use `lm` to run a regression of `mpg` on a few predictors in the data frame (choose two or three that you think would make a good model-- don't use all ten; we'll talk about why in later lectures).
Make sure to include `data = mtcars` as a keyword argument to `lm` so that R knows what data frame to use.

```{r}
lm_mtcars = lm(mpg ~ wt + hp + drat, data = mtcars)
preds = predict(lm_mtcars, mtcars)
residuals = mtcars$mpg - preds
ggplot(mtcars, aes(x = preds, y = residuals)) + geom_point() + geom_hline(yintercept = 0, linetype = "dashed", color = "red") + xlab("Fitted Values") + ylab("Residuals") + ggtitle("Residuals vs Fitted Values")
```

```{r}
plot(lm_mtcars,ask=F,which=1:2)
```

Briefly inspect the residuals plot by running `plot(lm.mtcars,ask=F,which=1:2)`.
What do you observe, and what does it mean?

***

The residuals plot shows no obvious patterns, which is desirable. This indicates that the linear regression model is a reasonable fit for the data.

***

### c) Interpreting the model

View the summary of your model by uncommenting and running the code below.
```{r}
summary(lm_mtcars)
```

Pick one of your predictors and give an interpretation of the estimate and standard error for its coefficient.
Be careful in your wording of the interpretation.

***

Let's consider the coefficient for 'wt' (weight). The estimate is -3.8778, and the standard error is 0.6328. This means that for each additional 1000 pounds of weight, the miles per gallon (mpg) is predicted to decrease by approximately 3.8778 units, with a standard error of 0.6328.

***

Which coefficients are statistically significantly different from zero? How do you know?

***

The coefficients for wt, hp, and drat are statistically significant, as their p-values are less than 0.05.

***

### d) Interpreting residuals

What is the Residual Standard Error (RSE) for this model? How many degrees of freedom does it have?

***

The Residual Standard Error (RSE) for this model is 2.561, with 28 degrees of freedom.

***

What is the value of $R^2$ for this model? (__Hint:__ look at the output of `summary`) Give an interpretation of this value.

***

The R^2 value for this model is 0.8369. This means that approximately 83.69% of the variation in mpg can be explained by the predictors wt, hp, and drat.

***

### e) Adjusted $R^2$

Briefly read about the adjusted $R^2$ [here](https://www.statisticshowto.com/probability-and-statistics/statistics-definitions/adjusted-r2/).
What is the adjusted $R^2$ of this model and how does this differ from the usual $R^2$ value? (__Hint:__ again, look at the output of `summary`).

***

The adjusted R^2 of this model is 0.8194, which is slightly lower than the usual R^2 value. The adjusted R^2 takes into account the number of predictors and the sample size, penalizing models with more predictors, unlike the regular R^2.

***

### f) CIs for coefficients

Read the documentation for the `confint` function, and use it to generate $95\%$ confidence intervals for the coefficients of your model.
Give an interpretation of these confidence intervals.

```{r}
confint(lm_mtcars)
```

***

The 95% confidence intervals for the coefficients are as follows:
(Intercept): [24.5278, 40.0239]
wt: [-5.2078, -2.5477]
hp: [-0.0458, -0.0066]
drat: [0.7262, 3.1483]
This means that we are 95% confident that the true value of each coefficient lies within the respective intervals. For example, we are 95% confident that the true value of the coefficient for 'wt' lies between -5.2078 and -2.5477.

***

## 4) the `cats` data set

The `cats` data set, included in the `MASS` library, contains data recorded from 144 cats.
Each row of the data set contains the body weight (`Bwt`, in kgs), heart weight (`Hwt`, in grams) and the sex (`Sex`, levels `'F'` and `'M'`) for one of the cats in the data set.

__Part a: plotting the data__

Create a scatter plot showing heart weight on the y-axis and body weight on the x-axis.
Ignore the `Sex` variable in this plot.

```{r}
library(MASS)
head(cats)
```

```{r}
ggplot(cats, aes(x = Bwt, y = Hwt)) + geom_point() + xlab("Body Weight (kg)") + ylab("Heart Weight (g)") + ggtitle("Heart Weight vs Body Weight")
```

Briefly describe what you see. Is there a clear trend in the data?

There is a clear positive linear trend in the data, indicating that as the body weight of the cats increases, so does the heart weight.

__Part b: fitting a linear model__

Fit a linear regression model to predict cat heart weight from cat body weight (and using an intercept term, of course).

```{r}
lm_cats = lm(Hwt ~ Bwt, data = cats)
summary(lm_cats)
```

Examine the coefficients of your fitted model.
What is the coefficient for the `Bwt` variable?
Interpret this coefficient-- a unit change in body weight yields how much change in heart weight?

***

The coefficient for the Bwt variable is 4.0341. This means that for each unit increase in body weight (kg), the heart weight (g) increases by 4.0341 units.

***

__Part c: back to plotting__

Create the same plot from Part a above, but this time color the points in the scatter plot according to the `Sex` variable.
You may use either `ggplot2` or the built-in R plotting tools, though I would recommend the former, for this.

You should see a clear pattern. Describe it. A sentence or two is fine here.

```{r}
ggplot(cats, aes(x = Bwt, y = Hwt, color = Sex)) + geom_point() + xlab("Body Weight (kg)") + ylab("Heart Weight (g)") + ggtitle("Heart Weight vs Body Weight (Colored by Sex)")
```

***

The plot shows that there is a clear separation between the sexes. Male cats generally have a higher body and heart weight compared to female cats.

***

__Part d: adding `Sex` and an interaction__

From looking at the data, it should be clear that the `Sex` variable has explanatory power in predicting heart weight, but it is also very correlated with body weight.

Fit a new linear regression model, still predicting heart weight, but this time including both body weight and sex as predictors *and* an interaction term between body weight and sex.
Take note of how R assigns `Sex` a dummy encoding.

```{r}
lm_cats_sex = lm(Hwt ~ Bwt + Sex + Bwt:Sex, data = cats)
summary(lm_cats_sex)
```

Examine the outputs of your model.
In particular, note the coefficients of `Sex` and the interaction between `Bwt` and `Sex`.
Are both of these coefficients statistically significantly different from zero?
How do you interpret the interaction term?

***

Both the SexM coefficient and the Bwt:SexM interaction term are statistically significantly different from zero (p-values are less than 0.05). The interaction term suggests that the effect of body weight on heart weight differs between male and female cats. Specifically, for each unit increase in body weight, the heart weight of male cats increases more than that of female cats.

***
